### 2018-03-08 Writing/disparity_review.tex
Minor edits to acknowledgemetns and addresses
Edits as far as methods

%Running head
\begin{flushright}
Version dated: \today
\end{flushright}
\bigskip
\noindent RH: disparate views on disparity.

@@

\noindent{\Large \bf Variations in multidimensional space.} 
\bigskip

@@

\noindent {\small \it 
$^1$School of Biological Sciences, University of Queensland, St. Lucia, Queensland, Australia.; $^2$NHM; $^{10}$Bristol}
\end{center}
\medskip
\noindent{+ These authors contributed equally to the manuscript.}
\noindent{*\bf Corresponding author.} \textit{guillert@tcd.ie}\\  
\vspace{1in}

@@

%1 -
% History of disparity methods
Since the time of the ancient, it has been observed that variation in morphologies often follows discrete patterns rather than a continuous gradient of forms.
These clusters of morphological groups, or, rather their differences (i.e. the disparity of shapes) was popularised in seminal palaeobiology papers in the 1990s [Runnegar 1987, Gould, Foote, Webb, etc.].
Disparity is since then used to describe a particular biological, phenotypic, and/or trait variation without strict distinction between the ontology and epistemology (see Semantic Box).
Over the last 30 years, disparity has been commonly used in the morphological disparity context.
The impressive amount of literature published since then tackles a vast array of biological questions, using different (sometimes contrasted) methods but often consists studies of the phenotype at a macroevolutionary level (i.e. species or above).
To this date, there is no consensus on whether disparity should be restricted and specific to these kind of studies or whether it should be used as a broader umbrella term designating studies of multivariate variation, not restricted in its questions, methods and disciplines.
Throughout this review, we aim to highlight the diversity of disparity analyses and how they can help understanding or provide a structure to biodiversity.
In essence, we suggest that disparity is a pattern of variation.

%2§ what use: commonly used
In evolutionary biology, the description of these patterns has been used to tackle a vast array of questions.
For example, it has been used (1) in Evo-devo to tackle questions of
%evo-devo
Developmental constraints [CITE];
Relationship between form and function and evolution functional traits [CITE] or
Scaling micro to macro [CITE];
% Ecology
(2) in ecology or palaeoecology to study:
intra-specific competition [CITE],
niche occupants, speciation and competition[CITE] or
macroecology (disparity in space)[CITE];
% Evolution
(3) on in macroevolution to study:
response to extinction and selectivity [CITE],
intra-specific competitive replacements [CITE] or
more globally, tempo and mode of evolution [CITE].

%3§ problems
Disparity studies have a great deal of promise in helping us to understand the evolution of biological diversity.
With the expansion of such methodological approaches in the last decade, however, we also identify three main issues that, if not tackled could reduce the power of such analysis.
First, many recent studies have lost focus; they are no longer hypothesis-driven, instead they are undertaken to characterise biological variation for its own sake.
Second, even where there is a clear question being explored, the data being used for disparity analyses is inappropriate and often driven by the availability of large datasets.
Third, the methods being used to analyse these datasets may not always be appropriate for the question and/or data at hand.
Here we deal with some of these issues and present best practice guidelines to allow researchers to optimise their questions before performing a disparity analysis.

\subsubsection{[Box]Disparity patterns or processes}
\label{box_semantic}
At the core of defining what \textit{is} disparity lies a potential divide between epistemology and ontology:
is there a difference between the pattern (and how researchers characterise it) and how it was ultimately generated?
If disparity describes patterns of variation, the underlying scientific question ultimately always needs to use them as a proxy for explaining the biological processes.
In the seminal papers from the late 80s, disparity was solely use to describe and understand the Cambrian Explosion were both the pattern (a high observed dissimilarity between groups) was directly linked to the process (and explosive radiation of multicellular life as championed by Gould).
Restricting disparity to this initial informal definition (multidimensional morphological at a macroevolutionary scale) can be advantageous for advances in this field but broadening it will allow more interdisciplinary discussion.
In fact, in the last years, disparity was also used as a potential tool for investigating ecological processes (disconnected from evolution: e.g. function [CITE], community structure[CITE], competition[CITE]).
In other words disparity could not only inform us on macroevolutionary patterns and processes but also something about environments and ecosystem functioning (with the advantages of easily allowing palaeocological and palaeoenvironemental perspectives).
\subsubsection{[Box]End}

@@

\section{Find a question}
Why to perform a disparity analysis?
This question might sound like an obvious question to start with but we would like to stress its crucial importance.
The last decade have seen a vast array of excellent methodological implementation and documentations rendering disparity analysis easy to perform \citep{bouxin2005ginkgo,oksanen2007vegan,geiger2008,zelditch2012geometric,adams2013geomorph,Claddis,dispRityv02,adams2017geometric}.
However, this facilitation might (conscientiously or not) pushes authors to perform disparity analysis ``just because'' (i.e. simply because of their facility rather than stemming from a conscious hypothetico-deductive approach).
It is thus important not to rush in a disparity analysis to avoid some basic statistical caveats.
For example, one major thing to consider is circularity: if ones aim is to explore the variation in the data set, a multi-dimensional analysis is perfectly valid.
However, the groups resulting from this exploratory analysis are then no more appropriate for testing a hypothesis [CITE stats 1.0].
Instead, it is more appropriate do the data exploration on a subset of the data (preliminary analysis) and then use the full data set to test the resulting hypothesis.
Clearly framing such hypothesis and their null expectancy will help designing better disparity analysis: which data and methods will be appropriate to test it?

\section{Collect the appropriate data}
If disparity is used to describe patterns for explaining some process, every possible data will always only be a proxy for these patterns.
Tangentially, more data is not necessarily better, especially when the data was originally designed and collected for different questions (e.g. cladistic data - see below).
Therefore, clearly framing the question as indicated above will help avoiding some major caveats.
For example, if a study is focussing on the ecological adaptive radiation of a group, ecological traits must be crucial an might be preferred to purely evolutionary traits.

\subsection{Which type of data}
Morphological disparity analysis can be based on three main types of data:
\begin{enumerate}
    \item discrete data that can be either describing feature (often referred to as ``cladistic''; e.g. the presence or absence of legs, the colour of a tail, etc. [CITE]) or more simply counts (e.g. number of digits, etc. [CITE]);
    \item continuous data that can be either direct measurements (e.g. body length, etc. [CITE]) or more sophisticated methods (e.g. geometric morphometrics, mathematical contour description, etc. [CITE]);
    \item or any mix and match of the two former (e.g. ratios [CITE]).
\end{enumerate}
None of these types of data are fundamentally better but rather each one will be more appropriate to a specific question/
For example, when looking at variation within bats species wings shape, homologous landmarks (for geometric morphometrics analysis) or continuous measurements on bones can be appropriate.
On the other hand, if the question focuses on the convergence between bats and birds wings, the homology of the traits to measure is more complex to define and the contour of the wings might more readily answer the question.

Regardless of the type of data, one really common follow up question is how much of it should be collected (both in terms of number of specimens and traits)?
\begin{itemize}
    \item \textit{Which taxonomic scope?}
    Whether disparity is a process or merely a method to describe a pattern (see Box \ref{box_semantic}) some studies have successfully use morphological disparity at both a inter-specific [CITE] or intra-specific level [CITE].
    The taxonomic scope will thus depends essentially on which type of variation ones want to analysis.
    For example, when studying skull shape plasticity in canids, it is essential to include intra-specific variation (specially domestic dogs) to capture the full scope of skull shapes.
    On the other hand, if the study focuses on canids skull variation through time, a inter-specific variation will not have major impact (up until the Holocene). %http://www.journals.uchicago.edu/doi/abs/10.1086/650372
    \item \textit{Which sample size?}
    The amount of data needed to test an hypothesis is by far not restricted to morphological disparity analysis.
    There is always a trade-of between the amount of data to collect and the time allocated to a project.
    In these conditions, it is mainly important to make sure that the data is collected fairly for subsequent statistical analysis.
    For example, if the question is to compare disparity among groups, one should collect, when possible, a similar number of elements per groups or.
    When this is not possible (e.g. for palaeontological data), one can always use statistical approaches to correct or asses biases due to data collection (e.g. rarefaction analysis [CITE]).
    \item \textit{Are the data consistent?}
    This is maybe the most important question related to the nature of the data.
    It is essential that the data is consistent and homogeneous to make sure that the disparity patterns can be linked to general processes rather than being simply the reflection of idiosyncrasies in the data set.
    For example, are the observed differences between two groups not related to the group's specificities (e.g. sexual dimorphism)?
    Are the elements fairly consistent and comparable them?
    Data can be missing, non overlapping, hierarchical (e.g. inapplicable), ambiguous, polymorphic or correlated.
    In any of these cases, are the patterns a mere reflection of these inconsistencies?
    For example, if ones use disparity analysis of locomotory modes among vertebrates and the collected data is hierarchical, limb data could create an artificial grouping of non-tetrapods and limbless tetrapods (e.g. snakes) that could not reflect processes.
\end{itemize}
In practice, however, these considerations can sometimes only be caveats to keep in mind.
The amount or type of data is often constrained by more mundane consideration like how much data can be collected in a particular space, time frame and financial set.

\subsection{Recycling data}
Following this more down to earth consideration, one common practice in disparity analysis is to recycle data from other studies.
This leads to another set of consideration on the source and the hypothesis tested with the original data.
One classic case is when using ``cladistic'' data in palaeobiology.
This data is usually collected for phylogenetic analysis and are thus designed to maximise differences between taxa to recover a clear phylogenetic signal.
This maximise difference can then artificially lead to maximised disparity among taxa [Foote - crinoids non-cladistic paper].
One relatively simple way to control for this specific bias can be to not use all the available data but select within the dataset the most pertinent characters for the question.
For example, in a study on disparity of dietary niches, removing non-dietary characters (e.g. sexually dimorphic ones) can be a good approach.
However, it is important to keep in mind that there is no one-to-one mapping between form a function and that some characters can be linked to multiple functions (e.g. limb traits can be both linked to locomotion and diet).
Ultimately, whether the data has been originally sourced or is recycled and whether any of the caveats described above are taken into account, disparity patterns will eventually be measured and analysed in another multitude of ways a we suggest that, whenever possible, this should first and foremost be considered in the design of the data collection protocol.

@@

@@

\section{Authors contribution}
TG, NC and PD proposed the idea of this review. TG, NC, SB, KD, AJ, SG, AG, KH, MH, GL, JOR, AP, ER, ES, ES, GS, GHT and PD wrote the manuscript,

>>>>

%Running head
\begin{flushright}
Version dated: \today
\end{flushright}
\bigskip
\noindent RH: disparate views on disparity. % NC: Needs to be changed?

@@

\noindent{\Large \bf Variations in multidimensional space.} 
% NC: Not sure about this title, especially now we are just doing a mini series. I think it really needs disparity in the title somewhere.
\bigskip

@@

\noindent {\small \it 
$^1$School of Biological Sciences, University of Queensland, St. Lucia, Queensland, Australia.\\ 
$^2$Department of Life Sciences, Natural History Museum, London, Cromwell Road, London, SW7 5BD, UK.\\ 
$^{10}$Bristol}
\end{center}
\medskip
\noindent{+ These authors contributed equally to the manuscript.}
\noindent{*\bf Corresponding author.} \textit{guillert@tcd.ie}\\  
\vspace{1in}

@@

%NC: OK I think this paragraph needs some rewriting. At present it's unclear to me what you're trying to tell me with this paragraph. Are you trying to tell me about the history of disparity? Are you trying to define it and tell me about the debates surrounding that? I think mostly you're aiming for the latter, but this needs to be much clearer. Try and be a bit more upfront about what you're doing, and don't be afraid of short, clear sentences without any fancy words. i.e. epistemiology and ontology helps no-one! Also note that you start by saying disparity = morphology, then you say it could be anything but people argue about it, then you say it can be anything. So perhaps you need to start broad and then say one type of disparity, and the most commonly analysed, is morphological, and that's what most people would think of when hearing the term. But it could apply to anything.

%1 -
% History of disparity methods
%Since the time of the ancient, it has been observed that variation in morphologies often follows discrete patterns rather than a continuous gradient of forms. % NC: Since the time of the ancients is super clunky. I've tried something different
Even classical scholars recognised that variation in the shape and form of organisms often follows discrete patterns rather than a continuous gradient.
% NC: changed the next bit as you need to define disparity/introduce it. As previously written it was hiding in a bracket.
Disparity analysis, the study of differences among these clusters of morphological groups, was popularised in seminal palaeobiology papers in the 1990s [Runnegar 1987, Gould, Foote, Webb, etc.].
Since then, disparity has been used to describe biological, phenotypic, and/or trait variation without strict distinction between the ontology and epistemology (see Semantics Box) 
% NC: Or glossary rather than semantics?
% NC: I have no idea what you mean with this ontology and epistemiology comment. 
Over the last 30 years, disparity has been most commonly used in the context of morphological disparity.
The impressive amount of literature published since then tackles a vast array of biological questions, using different (sometimes contrasted) methods but often consists studies of the phenotype at a macroevolutionary level (i.e. species or above).
To date, there is no consensus on whether disparity should be restricted and specific to these kind of studies or whether it should be used as a broader umbrella term designating studies of multivariate variation, not restricted in its questions, methods and disciplines.
Throughout this review, we aim to highlight the diversity of disparity analyses and how they can help understanding or provide a structure to biodiversity.
In essence, we suggest that disparity is a pattern of variation.

%2§ what use: commonly used
% NC: I feel like this perhaps need to fit in to the paragraph above, where you're saying it can be used for lots of things? 
The description of these patterns % NC: what patterns?
 has been used to tackle a vast array of questions.
For example, it has been used (1) in Evo-devo to tackle questions of
%evo-devo
Developmental constraints [CITE];
Relationship between form and function and evolution functional traits [CITE] or
Scaling micro to macro [CITE];
% Ecology
(2) in ecology or palaeoecology to study:
intra-specific competition [CITE],
niche occupants, speciation and competition[CITE] or
macroecology (disparity in space)[CITE];
% Evolution
(3) on in macroevolution to study:
response to extinction and selectivity [CITE],
intra-specific competitive replacements [CITE] or
more globally, tempo and mode of evolution [CITE].

%3§ problems %NC: Could maybe put the subheading "best practices" here?
Disparity studies have a great deal of promise in helping us to understand the evolution of biological diversity.
With the expansion of methodological approaches in the last decade, however, we also identify three main issues that, if not tackled could reduce the power of such analysis.
First, many recent studies have lost focus; they are no longer hypothesis-driven, instead they are undertaken to characterise biological variation for its own sake.
Second, even where there is a clear question being explored, the data being used for disparity analyses is inappropriate and often driven by the availability of large datasets.
Third, the methods being used to analyse these datasets may not always be appropriate for the question and/or data at hand.
Here we deal with some of these issues and present best practice guidelines to allow researchers to optimise their questions before performing a disparity analysis.

\subsubsection{[Box]Disparity patterns or processes}
\label{box_semantic}
At the core of defining what \textit{is} disparity lies a potential divide between epistemology and ontology: % NC: Can you either just write out what these terms mean, or add the explanation in brackets? I hate un-necessary jargon. Most PhD students will not be familiar with these terms. Were you before you got into this? (maybe you were, maybe the French education system is better!)
is there a difference between the pattern (and how researchers characterise it) and how it was ultimately generated?
If disparity describes patterns of variation, the underlying scientific question ultimately always needs to use them as a proxy for explaining the biological processes.
In the seminal papers from the late 80s, disparity was used to describe and understand the Cambrian explosion where both the pattern (high dissimilarities among groups) was directly linked to the process (an explosive radiation of multicellular life as championed by Gould REF).
Restricting disparity to this initial informal definition (multidimensional morphological dissimilarity at a macroevolutionary scale) can be advantageous for advances in this field but broadening it will allow more interdisciplinary discussion.
In recent years, disparity has been used to investigate ecological processes, disconnected from evolution (e.g. function [CITE], community structure[CITE], competition[CITE]).
Thus disparity % NC: I'm still unclear what you mean when you say disparity here. Do you mean the methods, the idea of discrete groups? I'm struggling a bit to parse how it can "inform us" per se.
could not only inform us on macroevolutionary patterns and processes but also something about environments and ecosystem functioning (with the advantages of easily allowing palaeocological and palaeoenvironemental perspectives).
\subsubsection{[Box]End}
% NC: I think I prefer the previous version where you actually had a subsection to discuss this within the text rather than a box. Seems particularly weird as you als odo some defining of disparity in the first paragraphs. Maybe move that into the box? Then just say here we define dispairty as "XXXX" in the text. Or shorten this and put it within the text? See what works. But the point of a box is to add details not needed in the text, so if it's in the text anyway then I'm dubious about the box...

@@

\section{Find a question} %NC: I wonder if "find a question" is exactly what we mean here. This suggests you've decided to use disparity whatever happens, and just need the right question. Maybe change to What is your question? 
Why are you performing a disparity analysis?
This seems like it should be obvious, but we stress its crucial importance.
The last decade has seen a vast array of new methods being published and implemented in freely available software, making disparity analyses easy to perform \citep{bouxin2005ginkgo,oksanen2007vegan,geiger2008,zelditch2012geometric,adams2013geomorph,Claddis,dispRityv02,adams2017geometric}.
However, this (consciously or not) pushes authors towards performing disparity analyses simply because of their ease of use, rather than stemming from a question driven approach.
Before rushing into a disparity analysis, one should consider some basic caveats.
For example, one major thing to consider is circularity: if the aim is to explore the variation in a dataset, disparity analysis is perfectly valid.
However, if this exploratory analysis reveals distinct groups, it would be circular to then test for differences among those groups as if this had been your hypothesis all along [CITE stats 1.0].
This can easily be avoided by using a subset of the data for preliminary data exploration, and then using the full dataset to test for differences among any groups that emerge.
Clearly framing your questions and their null expectations will help in designing better disparity analyses.

\section{Collect the appropriate data}
% If disparity is used to describe patterns that are explained by a certain process, every possible data will always only be a proxy for these patterns. % NC: Don't really understand your point here.
Once you have a clear question, it is important to collect the appropriate data for that question to avoid some major caveats.
For example, if a study focusses on the ecological adaptive radiation of a group, ecological traits are crucial and should be preferred over purely evolutionary traits.
Tangentially, more data are not necessarily better, especially when the data were originally collected for different questions (e.g. cladistic data - see below).

\subsection{Which types of data?} % NC: Not sure having these subheadings as Q looks best, maybe just Data types?
Morphological disparity analysis can be based on three main types of data:
\begin{enumerate}
    \item discrete data that can be either describing a feature (often referred to as ``cladistic''; e.g. the presence or absence of legs, the colour of a tail, etc. [CITE]) or counts (e.g. number of digits, etc. [CITE]);
    \item continuous data that can be either simple measurements (e.g. body length, etc. [CITE]) or more abstracted methods (e.g. geometric morphometrics, mathematical contour description, etc. [CITE]);
    \item or any mixture of discrete and continuous data (e.g. ratios [CITE]).
\end{enumerate}
None of these types of data are fundamentally better but each will be more appropriate for specific questions.
For example, when investigating variation within bat wing shapes, homologous landmarks (for geometric morphometrics analysis) or continuous measurements of bones may be appropriate.
On the other hand, if the question focuses on the convergence between bat and bird wings, the homology of the traits to measure is more complex to define and the contour of the wings might more readily answer the question. % NC: This section might need more explanation and/or some more examples

\subsection{How much data to collect?}
Regardless of the type of data, another important consideration is how much data to collect, in terms of numbers of both specimens and traits.
There are several important considerations.
\begin{itemize}
    \item \textit{Taxonomic scope}
    Morphological disparity has been studied at both inter-[CITE] and intra-specific levels [CITE].
    Taxonomic scope will depend on the question being investigated, and what type of variation is being analysed.
    For example, when studying skull shape plasticity in canids, it is essential to include intra-specific variation to capture the full variety of skull shapes, particularly that seen in domestic dogs (CITE).
    Conversely, if a study focuses on canid skull variation through time, interspecific data is most appropriate.
    % intra-specific variation will not have major impact (up until the Holocene). %http://www.journals.uchicago.edu/doi/abs/10.1086/650372
    \item \textit{Sample size}
    There is always a trade-off between collecting enough data and the time allocated to a project.
    It is important to ensure that the data is collected in a way that will not bias subsequent statistical analysis.
    For example, if the aim is to compare disparity among groups, one should collect, where possible, a similar amount of data for each group.
    When this is not possible (e.g. for palaeontological data), one should use statistical approaches to correct or assess biases due to data collection (e.g. rarefaction analysis [CITE]).
    Equally, one should ensure that enough data is available so that the methods have power to detect the patterns being tested.
    \item \textit{Are the data consistent?} %NC: Not suer I get this section. What do you mean by consistent???? Also does this fit into the "how much data" section or is it a seaprate consideration?
    This is maybe the most important question related to the nature of the data.
    It is essential that the data is consistent and homogeneous to make sure that the disparity patterns can be linked to general processes rather than being simply the reflection of idiosyncrasies in the data set.
    For example, are the observed differences between two groups not related to the group's specificities (e.g. sexual dimorphism)?
    Are the elements fairly consistent and comparable them?
    Data can be missing, non overlapping, hierarchical (e.g. inapplicable), ambiguous, polymorphic or correlated.
    In any of these cases, are the patterns a mere reflection of these inconsistencies?
    For example, if one uses disparity analysis of locomotory modes among vertebrates and the collected data is hierarchical, limb data could create an artificial grouping of non-tetrapods and limbless tetrapods (e.g. snakes) that could not reflect processes.
\end{itemize}
Of course in practice some of these caveats are unavoidable. 
Data collection is often constrained by the time available, financial considerations and other unavoidable issues, however, these caveats should be kept in mind during the analyses and subsequent interpretation of the results. % NC: I'd be tempted to move this to the end of the section to finish it off.

\subsection{Recycling data}
A common practice in disparity analysis is to recycle data from other studies which can introduce additional problems.
One classic case is using ``cladistic'' data in palaeobiology.
These data are usually collected for phylogenetic analysis and are thus designed to maximise differences among taxa to recover a clear phylogenetic signal.
This data can therefore artificially lead to maximised disparity among taxa [Foote - crinoids non-cladistic paper].
A relatively simple way to reduce this specific bias is to select the most pertinent characters for the question from the dataset, rather than using all the data.
For example, in a study of disparity of dietary niches, removing non-dietary characters is a good approach.
However, it is important to remember that there is no one-to-one mapping between form and function, and that some characters can have multiple functions (e.g. limb traits can be used for both locomotion and feeding).
%Ultimately, whether the data was collected for the purpose of the study or is recycled, and whether any of the caveats described above are taken into account, disparity patterns will eventually be measured and analysed in another multitude of ways a we suggest that, whenever possible, this should first and foremost be considered in the design of the data collection protocol. 
%NC: Not sure these linking statements are needed

@@

% NC: I do like this quote. The actual quote is: "Your scientists were so preoccupied with whether or not they could that they didn't stop to think if they should." Dr Ian Malcolm. It would be magnificent ot get a jurassic park quote into a paper...

@@

\section{Authors contribution}
TG, NC and PD proposed the idea of this review. All authors edited the manuscript.
